---
title: "Josh_Kemp_BCB_546_R_assignment.md"
author: "Joshua Kemp"
date: "2023-03-10"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Joshua Kemp R Assignment

### Prepare Environment

#install.packages("tidyverse", "dplyr", "sjmisc") \# remove \# if any are needed #install.packages("sjmisc")

```{r libraries, include=FALSE}
    library(tidyverse)
    library(dplyr)
    library(rmarkdown)
    library(sjmisc)
    library(stats)
```

## Download files and transpose fang

```{r download and transpose, include=FALSE}
    snppositions <- read_tsv("https://github.com/EEOB-BioData/BCB546_Spring2023/raw/main/assignments/UNIX_Assignment/snp_position.txt", col_names=TRUE, show_col_types = FALSE) %>% select(SNP_ID, Chromosome, Position)
    fang.original<- read_tsv("https://github.com/EEOB-BioData/BCB546_Spring2023/blob/main/assignments/UNIX_Assignment/fang_et_al_genotypes.txt?raw=true", col_names=FALSE, show_col_types = FALSE)
```

#Col.names = FALSE makes sure that the header will be a row so that we have the SNP column when we transpose. I had added the column using mutate, but this is cleaner.

### Data Inspection

There are genotypes in the fang file that do not belong to teosinte or maize groups we filter for looks like they come from the related group

#R studio is nice, because it is easy to open a file or element and take a look.


```{r File Inspection, include=TRUE}
    unique(duplicated(fang.original$X1)) #make sure these are non-duplicated
    unique(fang.original$X3) # see what is here, shows that we are leaving out some of the groups present.  I assume they belong to trypsicum
    unique(duplicated(snppositions$SNP_ID)) # just making sure that these are unique, but since the unix project had no problems, they should be
    str_which(unique(snppositions$Position), "\\D")  #  Check for non-digit values in position. This block looks for values in postion that could be problems.  Found multiple, unknown, and NA
      str_which(unique(snppositions$Position), "\\s") # check for whitespace
      Filter(is.null,unique(snppositions$Position)) # look for nulls
      Filter(is.na,unique(snppositions$Position)) # look for NAs
      unique(snppositions$Position)[55] #what non-digit thing is element 55 and 89
      unique(snppositions$Position)[189]
```

### Separate by species and transpose

```{r Seperate Fang Genotypes by Species, include=FALSE}
    maize.fang <- filter(fang.original, X3 == "ZMMIL" | X3 == "ZMMLR" | X3 =="ZMMMR" | X1 == "Sample_ID") %>% rotate_df()  #pull out Maize groups, write to a file, and transpose
    teosinte.fang <- filter(fang.original, X3 == "ZMPBA" | X3 == "ZMPIL" | X3 =="ZMPJA" | X1 == "Sample_ID") %>% rotate_df()  #pull out teosinte groups, write to a file, and transpose
```


## Functions and Lists

#### Lists

```{r create lists, include=FALSE}
    species.files.names <- c("maize.fang", "teosinte.fang")  # list of starting files
    chromosomes <- c(1:10)   # list of Chromosome values
    problem.locations <- c("unknown", "multiple", NA, "missing")  # None numeric values present in the position column
```

#### Functions

```{r create functions, include=FALSE}
    Firstrow.to.colnames <- function(dfx) {
      newcolnames <- filter(dfx, V1 == "Sample_ID")
      colnames(dfx) <- c(newcolnames)
      dfx }  ## filter the row containing header information and assign column names based on that vector
    Merge.with.snppositions <- function(f) merge.data.frame(snppositions, f, by.x = "SNP_ID", by.y = "Sample_ID", all.y = TRUE, sort = FALSE) # Join a data frame with the snppositions file, keeping the extra rows (which contain some useful information)
    Remove.problem.postions <- function(dfx) postions.problems <- filter(dfx, Position %in% problem.locations)  #find and filter all rows from a dataframe contatining positions the non numeric positions in our list and write them to a new seperate dataframe called problem positions.
    Replace_quest_with_hyphen <- function(dfx) as.data.frame(lapply(dfx, function(x) (gsub(pattern="?", replacement="-", x, fixed = TRUE))) -> dfx) # Replace all questionmarks with hyphens within a dataframe
    Filter.by.chromosome <- function(dfx) {
      append(
      lapply(chromosomes, function(y) filter(dfx, Chromosome == y )) %>% lapply(function(x) filter(x, !(Position %in% problem.locations)) %>% arrange(as.numeric(Position))),
      lapply(chromosomes, function(y) filter(Replace_quest_with_hyphen(dfx), Chromosome == y )) %>% lapply(function(x) filter(x, !(Position %in% problem.locations)) %>% arrange(desc(as.numeric(Position))))
      )
      } # This is complicated, but it the lapply version of a nested loop.  It takes a given data frame, and for each chromosome within our chromsosme list, it will filter by chromosome, remove problem locations and then sort. I does this twice once ascending, and onnce decending on a version of the dataframe with the questionmarks replaced.
    Create.Split.Filenames <- {
      sapply(1: length(species.files.names), function(x) {
      c(
        sapply(chromosomes, function(y) c(paste0(species.files.names[x], "_chr-", y, "_ascending"))),
        sapply(chromosomes, function(y) c(paste0(species.files.names[x], "_chr-", y, "_descending"))))
      })
      } # Does the samething basically, but just modifying the file names so we can assign them later
    Write_csv <- function(df) walk2(df, paste0(names(df), ".csv"), write_csv) # writes a dataframe to file in the current working directory
```

## Workflow

#### Add Names and Marker locations (Chromosome and Postion) to our genotype files then name them

```{r Run Data processing, include=FALSE}
    genotypes.by.species <- list(maize.fang, teosinte.fang)
    genotypes.by.species <- lapply(genotypes.by.species, Firstrow.to.colnames)
    merged.genotypes.by.species <-lapply(genotypes.by.species, Merge.with.snppositions)
    names(merged.genotypes.by.species) <- c(paste("merged_", species.files.names, sep = ""))
    #### Remove problematic marker positions
    problem.positions <- lapply(merged.genotypes.by.species, Remove.problem.postions)
    names(problem.positions) <- paste0(species.files.names, "_miss_unk_mult", sep = "")
    problem.positions_hyphen <- (lapply(problem.positions, function(l) Replace_quest_with_hyphen(l)))
    names(problem.positions_hyphen) <- paste0(species.files.names, "_hypen_miss_unk_mult", sep = "")
    #### Split dataframes by Chromosome
    split.genotype.files <- lapply(merged.genotypes.by.species, Filter.by.chromosome)
    Create.Split.Filenames -> newsplitnames
    #### Flatten List Structure (remove nesting) and Name the split files
    flat.split.genos <- list_flatten(split.genotype.files)
    names(flat.split.genos) <- newsplitnames
    #### write lists to csv files
    Write_csv(flat.split.genos)
    Write_csv(problem.positions)
    Write_csv(problem.positions_hyphen)
```

## Visualization

### Reformatting

```{r Reformat, include=FALSE}
    both_species.dataframe <- full_join(merged.genotypes.by.species[[1]], merged.genotypes.by.species[[2]], by = join_by("SNP_ID" == "SNP_ID", "Chromosome" == "Chromosome", "Position" == "Position"), suffix = c("", ""))
    transposed_full.dataframe <- rotate_df(both_species.dataframe)
    transposed_full.dataframe <- rotate_df(both_species.dataframe) %>% mutate(Species = ifelse(V986 == "ZMMIL" | V986 == "ZMMLR" | V986 =="ZMMMR", "Maize", ifelse(V986 == "ZMPBA" | V986 == "ZMPIL" | V986 =="ZMPJA", "Teosinte", NA)), .before = V1)
    transposed_full.dataframe[1,1] <- "Species"
    colnames(transposed_full.dataframe) <- filter(transposed_full.dataframe, Species == "Species")
    reformatted_one_genotypecall_per_row.df <- filter(transposed_full.dataframe, !(Sample_ID == "NA" | Sample_ID == "Sample_ID")) %>% pivot_longer(cols = -c("Species", "Group", "JG_OTU", "Sample_ID")) %>% 
      merge.data.frame(snppositions, by.x = "name", by.y = "SNP_ID", all.x = TRUE, sort = FALSE) %>%    # quick way to add snp locations
      relocate(c(Chromosome, Position, Species, Group), .before = name) %>%   #move these columns to the front of the df
      mutate(Allele1 = substr(value, 1, 1), .before = value) %>%    #take first allele call, difit 1, and create a column
      mutate(Allele2 = substr(value, 3, 3), .before = value) %>%    #take 3rd digit which is the second allele call
      mutate(Homozygous = ifelse(Allele1 == "?" | Allele2 == "?", NA, eval(Allele1 == Allele2)), .before = Allele1) %>%  #if there is no call "?", return NA otherwise, check if allele 1 equals allele 2 and return true or false
      mutate(Missing_call = Allele1 == "?" | Allele2 == "?", .before = Allele1) %>% # if allele one or two is a missing call write True, otherwise false and write to new column called Missing_call
      filter(!(Position %in% problem.locations)) %>% 
      mutate(bin_position = cut(as.integer(Position), 20, labels = FALSE), .after = Position)
```

## Plotting

#### SNPs per Chromosome

```{r plot SNPs Per Chromosome, include=TRUE}

    ggplot(data = reformatted_one_genotypecall_per_row.df ) + geom_bar(mapping = aes(x= as.factor(as.integer(Chromosome)), color = Species )) +
      ylab('Number of tested SNPs') +
      xlab('Chromosome') 
      
```

#### Distribution across Chromosomes

```{r plot snp density fuction by position, include=TRUE}
    ggplot(data = reformatted_one_genotypecall_per_row.df, aes(as.double(Position))) +
      geom_density() +
      scale_color_viridis_d() +
      facet_wrap(~ as.double(Chromosome))
```

#### Homozygous calls across Chromosome Positions

```{r setup, include=TRUE}
    ggplot(data = reformatted_one_genotypecall_per_row.df, aes(x = as.double(Position), fill = (Homozygous), position = "fill", color = Species)) +
      geom_density(alpha = .2) +
      facet_wrap(~ as.double(Chromosome))
```


#### Missing_call by Position


```{r summarise missing, include=FALSE}
    contaminated_lines <- summarise(reformatted_one_genotypecall_per_row.df, Mean_Missing = mean(Homozygous, na.rm = TRUE), .by = c(Sample_ID, Species))

```

```{r plot missing by position, include=TRUE}
    ggplot(data = reformatted_one_genotypecall_per_row.df, aes(x = as.double(Position), fill = Missing_call, position = "fill", color = Species)) +
      geom_density(alpha = .2) +
      scale_color_viridis_d() +
      facet_wrap(~ as.double(Chromosome))
```

##### Look at how inbred our lines in our samples are, which could indicate problems and lines that may need to be omitted

```{r plot homozygous by sampleID, include=TRUE}
    contaminated_lines <- summarise(reformatted_one_genotypecall_per_row.df, Mean_Homozygosity = mean(Homozygous, na.rm = TRUE), .by = c(Sample_ID, Species))
    contaminated_lines <- arrange(contaminated_lines, Mean_Homozygosity)
    head(contaminated_lines, n = 20)
```

```{r plot homozygousity by species, include=TRUE}
    ggplot(data = contaminated_lines) +
      geom_point(mapping = aes(x= Species, y= Mean_Homozygosity, color = Species, alpha=0.05)) +
      ggtitle("Sample Homozygosity by Species")
```


My machine runs out of memory running this 

```{r sample homozygousity may crash computer, include=TRUE}
    ggplot(data = contaminated_lines) + geom_point(mapping = aes(x= Species, y= Mean_Homozygosity, color = Species, alpha=0.05)) + ggtitle("Sample Homozygosity by Species")
```

Many of these lines are less inbred than I would expect, not sure if hybrids or complex lines were included on purpose. Depends on the type of study this was for.


#### Regions with higher or lower homozygosity
```{r summarise homozygousity by position, include=FALSE}
    Homozygosity <- summarise(reformatted_one_genotypecall_per_row.df, Mean_Homozygosity = mean(Homozygous, na.rm = TRUE), .by = c(Chromosome, Position, Species))
```

```{r plot homozygousity by position, include=TRUE}
    ggplot(data = Homozygosity, aes(x = as.double(Position), y = (Mean_Homozygosity), color = Species)) +
      geom_point(alpha = .6) +
      facet_wrap(~ as.double(Chromosome))
```

#### Samples with Missing data

```{r summarise missing, include=TRUE}
    Missing <- summarise(reformatted_one_genotypecall_per_row.df, Mean_Missing = mean(Missing_call, na.rm = TRUE), .by = c(Sample_ID, Species))
    Missing <- arrange(Missing, desc(Mean_Missing))
    head(Missing, n = 20)
```

```{r plot missing, include=TRUE}
    ggplot(data = Missing, aes(x = Sample_ID, y = Mean_Missing, color = Species)) +
      geom_point(alpha = .6)
````



